---
title: "PMA: A Brief Look at Facebook Posts"
author: "Alexander Hoyle"
date: "December 27, 2016"
output:
  html_document:
    code_folding: show
---

## Motivation

Although the impending analysis will use news sources' text data we collected in the
last post, I've decided to take a little detour and take a look at the Facebook
pages that supported the article collection. I hope this will give us some broad
insight into the nature of our data and the divided media ecosystem on Facebook.

For this task, I'm using R, because I find it (combined with 
[`dplyr`](https://github.com/hadley/dplyr)
and [`magrittr`](https://github.com/tidyverse/magrittr))
more suited to readable, procedural analysis of rectangular data than Pandas is.
I also prefer `ggplot` to `matplotlib`.

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(magrittr)
library(ggplot2)
library(extrafont)

apply_plot_theme <- function(){
  ## apply a common theme to a plot
    theme_bw() +
    theme(text = element_text(family = 'Calibri'))
}
```

## Getting the data

First, we'll pull the article information we stored in the database.

I'm also merging in the Facebook study data on source partisanship in order to 
incorporate source alignment into our analysis.

```{r collect db data, eval=FALSE}
article_db <- src_postgres('articles', user = 'postgres', password = 'postgres')
fb_posts <- tbl(article_db, 'fb_posts')

post_data <- 
  fb_posts %>%
  select(post_id, page_id, created_time, shares, retrieved_on) %>%
  collect(n = Inf) %>% # dplyr is lazy unless explicitly told otherwise
  filter(!duplicated(post_id)) %>%
  
  # create date columns
  mutate(post_date = as.Date(created_time),
         retrieved_date = as.Date(retrieved_on, '%a %b %d %H:%M:%S %Y'),
         post_month = lubridate::floor_date(post_date, 'month'),
         post_week =  lubridate::floor_date(post_date, 'week'))
```

```{r read, echo=FALSE}
readRDS('C:/post_data.rds')->post_data
```

```{r collect external data, warning=FALSE, message=FALSE}
alignment_data <-
  read.csv('https://raw.githubusercontent.com/jonkeegan/blue-feed-red-feed-sources/master/included_sources.csv',
           stringsAsFactors = FALSE) %>%
  mutate(page_id = as.character(fb_id)) %>%
  select(-fan_count)

#updated fan counts gatherered using the FB graph API
fan_counts <-
  read.csv('./input/fan_count_20161228.csv',
           colClasses = c('numeric', 'character'))
  
alignment_data %<>% inner_join(fan_counts)
```

## Posts over Time

Just to get a sense of our data, we'll plot the number of posts and shares below.
I look at distributions over the week to smooth the data and remove any cyclical
effects.

```{r plot function}
# important dates
sig_dates <-
  data_frame(date = as.Date(c('2016-07-21', '2016-07-28', '2016-09-26',
                              '2016-10-04', '2016-10-19', '2016-11-08')),
             event = c('Rep. Convention', 'Dem. Convention', '1st Debate',
                       '2nd Debate', '3rd Debate', 'Election Day'))

plot_post_data <- function(data, ydat, ylab, ymax, plot_title, sig_dates){
  ysteps <- seq(from = ymax, by = -ymax / 18, length.out = nrow(sig_dates))
  
  data %>%
    # plot
    ggplot(aes_string(x = 'post_week', y = ydat)) +
    stat_summary(geom = 'smooth', color = '#00467f',
                 fun.ymin = function(x) quantile(x, 0.05),
                 fun.y = mean,
                 fun.ymax = function(x) quantile(x, 0.95)) +
    
    # add significant dates to plot
    geom_vline(data = sig_dates, 
               aes(xintercept = as.numeric(lubridate::floor_date(date, 'week'))),
               color = '#6A7277', alpha = 0.7) +
    geom_label(data = sig_dates,
               aes(x = lubridate::floor_date(date, 'week'),
                   y = ysteps,
                   label = event),
               hjust = -0.015, family = 'Calibri', size = 3.5,
               alpha = 0.65,
               label.padding = unit(0.15, 'lines'),
               label.size = NA) +
    
    # labels
    labs(title = plot_title,
         subtitle = '(bands show 90% interval, central line is mean)',
         x = 'Date (Weekly)',
         y = ylab) +
    scale_y_continuous(labels = scales::comma) +
  
    # theming
    apply_plot_theme()
}

```

Note the increase in shares per post around the first debate and
the subsequent dropoff following the election (most evident at the 95% percentile
than at the mean).

```{r plot posts, fig.width=8}
# plot weekly post data
post_data %>%
    inner_join(alignment_data, by = 'page_id') %>%
    filter(post_date >= as.Date('2015-06-15'),
           post_date < as.Date('2016-12-25')) %>%
    group_by(post_date, post_week, page_id) %>%
    summarize(posts = n()) %>%
    plot_post_data(ydat = 'posts', ylab = 'Posts per Page', ymax = 40,
                   plot_title = 'Distribution of Posts per Page',
                   sig_dates)

# plot weekly share data
post_data %>%
    filter(!is.na(shares),
           post_date >= as.Date('2015-06-15'),
           post_date < as.Date('2016-12-25')) %>%
    plot_post_data(ydat = 'shares', ylab = 'Shares', ymax = 4500,
                   plot_title = 'Distribution of Shares per Post',
                   sig_dates)
    

```

## Effect of Partisanship on Sharing

I thought it might be worthwhile to determine if there exists a relationship
between the level of page partisanship and the likelihood of sharing a given
post from that page.

To this end, I group the post data by source and merge in the Facebook page
alignment data. Here, pages are assigned a value on a [-2, +2] scale from left-leaning
to right-leaning. I take its absolute value to create a measure of partisanship
irrespective of political alignment.

I then regress the number of shares per post on this measure, controlling for total
followers and whether the source is left- or right-aligned. (Note that I remove a
low-partisan point -- the *Tampa Bay Times* at 0.55 on our 0 to 2 scale -- since it has
high leverage).

We see that, on the left side of the spectrum, source partisanship appears to have
a modestly significant relationship with the average number of shares per post.
Since 'partisanship' is based  on the distribution of follower political identification,
this supports the idea that those who are most ideologically driven are also most
prone to sharing ("shares" being the [most important measure](http://www.adweek.com/news/advertising-branding/brands-favor-social-shares-over-likes-148256)
of engagment on Facebook).

Using post-level data would permit us to more meaningfully comment on the relationship between 
partisan content and shares (controlling for sharer ideology), but
we don't have such granular measures of bias. Rather, we are limited to the level of
the page generating each post (I suppose we could use bootstrapping to estimate variance,
but I think the below finding is sufficiently interesting for now). It's possible
we'll develop a useful measure of partisanship based on the text data in a future
post, at which point we can return to this question.

As an aside, we are only looking at a subset of the `r n_distinct(post_data$page_id)`
most partisan (and most popular) sources of the 500 found in the [Facebook study on ideological news](https://dataverse.harvard.edu/file.xhtml?fileId=2676555&version=2.2)
that I cited in my first post (this is a result of filtering performed by the *Wall
Street Journal* in the data they made public for 
["Blue Feed, Red Feed"](http://graphics.wsj.com/blue-feed-red-feed/)). I've requested
access to the data and will run this regression on the full set of sources at a later date.

```{r regression on pages}
page_data <-
    post_data %>%
    filter(post_date >= as.Date('2015-06-15'),
           post_date <= as.Date('2016-12-25')) %>%
    group_by(page_id) %>%
    summarize(shares_per_post = mean(shares, na.rm = TRUE)) %>%

    # include alignment data
    inner_join(alignment_data, by = 'page_id') %>%
    mutate(partisanship = abs(avg_align),
           side = factor(side)) %>%
    filter(partisanship > 0.55)

model <-
  lm(log(shares_per_post) ~ side:partisanship + log(fan_count) + side,
     data = page_data)

pander::pander(model)
```

Here's a plot that roughly approximates the regression.

```{r plot regression like data}
page_data %>%
  # plot
  ggplot(aes(y = log(shares_per_post) / log(fan_count), x = partisanship, color = side)) +
  geom_point() +
  geom_smooth(method=lm) +
  
  # labels
  labs(title = 'Follower-Adjusted Shares by Partisan Alignment of News Source',
       x = 'Source Partisanship',
       y = 'Log(Shares Per Post) / Log(Total Followers)') +
  
  # themeing
  scale_color_manual(name = 'Side', values = c('#00467f', '#EF4623')) +
  apply_plot_theme()
```